---
title: Static plots - Assignments
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 5)
```

### References and datasets

-   Original paper: Abbas et al. (2019) *Ecological Indicators* (<https://doi.org/10.1016/j.ecolind.2019.105524>)
-   StatQuest: PCA and other videos (<https://statquest.org/>)
-   Also check the lecture slides for L02-L07

# 1. Student ID and Name

## [Student ID]{style="color:red"}: 20957364

## [Name]{style="color:red"}: YIP Chi Ho

# 2. Generative AI declaration

Students may use generative AI tools during assignment, for example, to check and improve code, provided that the content is original and reflects the student's own contributions. If generative AI is used, students must declare its use in the box below. An example statement is: “*During the assignment, I used [XXXX] to [XXXX]. I subsequently reviewed and edited the code and take(s) full responsibility for the final version of the code.”*

During this assignment, I used generative AI to write the code that combine multiple data into a data frame, and improve code quality. I subsequently reviewed and edited the code and take(s) full responsibility for the final version of the code.

# 3. Assignments

## Github: Create a Github account

-   Create your Github account.
-   Install Git on your computer as well (the latest version = 2.53.0).
-   Not mandatory, but please read the advice about usernames in the lecture.

## Github: Create your first repository

-   Repository name must be “DASC3240_Assignment1.”
-   Make the visibility “public” so I can check it.
-   The repository must have a README and a License file.
-   Include a screenshot of your shell window showing the version of Git on your computer.

## Tree community data in Hong Kong

Go to Canvas and download `Abbas_2019_TableS4_ed1.csv` and `Abbas_2019_TableS4_ed2.csv`.

-   `Abbas_2019_TableS4_ed1.csv` includes the plot information (metadata).
-   `Abbas_2019_TableS4_ed2.csv` includes dominant tree species data in the 28 forest plots.

1.  Calculate the total abundance of the dominant tree species in each forest plot, and visualize the relationship between the total abundance and forest age. You should use `AGE_MEDIAN` column in the `Abbas_2019_TableS4_ed1.csv` file as a forest age index.

2.  Visualize the dominant tree species compositions using dimension reduction techniques. For visualization, you must use different colors for different forest age groups.

    2-1. Perform PCA to visualize the tree species composition.\
    2-2. Perform NMDS to visualize the tree species composition.\

# 3. Marking criteria and important notes

[**Please check the lecture PDF "L08_Assignment-I.pdf" provided on Canvas.**]{style="color:darkred"}

# 4. Your answers

### 1. Calculate the total abundance of the dominant tree species in each forest plot, and visualize the relationship between the total abundance and forest age.

-   You should use `AGE_MEDIAN` column in the `Abbas_2019_TableS4_ed1.csv` file as a forest age index.
-   You can calculate **the sum of each column** in `Abbas_2019_TableS4_ed2.csv` to calculate the total abundance of the dominant tree species in each plot.
-   You will combine the data from two dataframes, so please check if the column names of `Abbas_2019_TableS4_ed2.csv` and the row names (indices) of `Abbas_2019_TableS4_ed1.csv` match (by using R script, not by visual inspection!).
-   To visualize the relationship, barplot is **NOT** recommended.

```{r q1-plot}
library(tidyverse)

# load dataset
ed1_df <- read.csv('data/Abbas_2019_TableS4_ed1.csv', row.names = 1)
ed2_df <- read.csv('data/Abbas_2019_TableS4_ed2.csv', row.names = 1)

# check if the row names of ed1_df match the column names of ed2_df (result: TRUE)
all(rownames(ed1_df) == colnames(ed2_df))

# age median in each forest plot
age_median <- ed1_df[['AGE_MEDIAN']]

# total abundance of the domiant species in each forest plot
total_abund <- colSums(ed2_df)

# build a data frame of the total abundance and the age median of each forest
sca_df <- data.frame(
  TotalAbund = total_abund, 
  AgeMedian = age_median
)

# plot the sca_df (scatter plot data frame)
ggplot(sca_df, aes(x = factor(AgeMedian), y = TotalAbund)) + # factor(AgeMedian) allows box plot in each age group
  geom_boxplot() + 
  geom_jitter(width = 0.2, size = 2, alpha = 0.7) + 

  labs(
    title = "Total abundance of dominant tree species against forest age",
    x = "Forest Age Median (years)",
    y = "Total abundance of domiant tree species (individuals)"
  )



```

### 2. Visualize the dominant tree species compositions using PCA and NMDS.

There are a number of points you should consider:

-   For visualization, you must use different colors for different forest age groups. Use of different marker styles for different groups is also a good idea.
-   [**Set random seeds if your analysis includes random process(es)**]{style="color:darkred"}.
-   Before applying dimension reduction functions (e.g., `prcomp()`), please make sure that [**rows represent samples (= forest plots) and columns represent tree species**]{style="color:darkred"}. You have to transpose the data frame (In `Abbas_2019_TableS4_ed2.csv`, columns represent samples and rows represent tree species).
-   Usually we should consider whether we must perform the data preprocessing before applying dimension reduction techniques (e.g., standardization). [**In this assignment, please convert your data into compositional data using the following code**]{style="color:darkred"} `d_tree_rel <- apply(d_tree, 2, function(x) x/sum(x))` (`d_tree` represents the name of your data.frame. You should change it depending on your R object name). This means that, in this assignment, we are interested in tree community compositions, not abundances.

#### 2-1. Perform PCA to visualize the tree species composition.

After PCA, please visualize the result.

```{r pca-plot}
library(tidyverse)

d_tree_rel <- apply(ed2_df, 2, function(x) x / sum(x))

# d_tree_t is the transpose of d_tree_rel
d_tree_t <- t(d_tree_rel)

# Do PCA on d_tree_t
pca_result <- prcomp(d_tree_t, center = FALSE, scale. = FALSE)

# Get variance of pca, pc1 and pc2
pca_var <- summary(pca_result)$importance
pc1_var <- round(pca_var[2, 1] * 100, 1)
pc2_var <- round(pca_var[2, 2] * 100, 1)

# Cast the pca result into a data frame with the age group
pca_df <- as.data.frame(pca_result$x)
pca_df$AgeGroup <- factor(ed1_df[rownames(pca_result$x), "AGE_MEDIAN"])

# Plot PCA result with colors by age group
pca_df %>%
  ggplot(aes(x = PC1, y = PC2, color = AgeGroup, shape = AgeGroup)) +
  geom_point(size = 3) +
  labs(
    title = "PCA of Tree Species Composition",
    x     = paste0("PC1 (", pc1_var, "%)"),
    y     = paste0("PC2 (", pc2_var, "%)"),
    color = "Age Median (years)",
    shape = "Age Median (years)"
  ) +
  theme_bw()
```

#### 2-2. Perform NMDS to visualize the tree species composition.

After NMDS, please visualize the result. You can use Bray-Curtis dissimilarity to calculate the dissimilarity among plots.

```{r nmds-plot}
library(vegan)

# Set the random seed for NMDS
set.seed(3240)

# Run NMDS with Bray-Curtis dissimilarity
nmds_res <- metaMDS(d_tree_t,
                    distance = "bray",
                    k = 2) # dimension = 2

# Cast the NMDS result into a data frame and combine with the age group
nmds_res_df <- as.data.frame(nmds_res$points)
nmds_res_df$AgeGroup <- factor(ed1_df[rownames(nmds_res$points), "AGE_MEDIAN"])

# Plot NMDS result with colors by age group
nmds_res_df %>%
  ggplot(aes(x = MDS1, y = MDS2, color = AgeGroup, shape = AgeGroup)) +
  geom_point(size = 3) +
  labs(
    title  = "NMDS of Tree Species Composition (Bray-Curtis)",
    x      = "Axis 1",
    y      = "Axis 2",
    color  = "Age Median (years)",
    shape  = "Age Median (years)"
  ) +
  theme_bw()
```
